---
output:
  md_document:
    variant: markdown_github
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
stopifnot(require(knitr))
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = TRUE,
  fig.asp = 0.8,
  fig.width = 4,
#  out.width = "60%",
  fig.align = "center")
library(rstan)
library(sf)
rstan_options(auto_write = TRUE)
```

# geostan 

## Bayesian Spatial Analysis

### Overview

The **geostan** R package provides a user-friendly interface to hierarchical Bayesian models (HBMs) for areal data. It is designed for ease of use and promotion of sound spatial analysis workflows with a particular emphasis on spatial epidemiology and survey data. All of the models are fit using the Stan probabilistic programming language. The package is still under development.

The following models are available:

* Generalized linear models with Gaussian, Student’s *t*, Poisson, and Binomial likelihood functions.
* Eigenvector spatial filtering (ESF) models.
* Intrinsic conditional autoregressive (IAR) models.
* The Besag-York-Mollie (BYM2) model introduced by Riebler et al.

All of the models are able to incorporate additional 'varying intercept' terms for partial pooling of observations, and can also model data uncertainty using, e.g., standard errors on estimates from the American Community Suvey.

### Package installation

You can try out a development version of the package. Since it is not yet on CRAN, unmet dependencies must be installed first. You should first install Rstan by following the [getting started guide](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started). Then you'll also need the following R packages:

```{r eval=FALSE}
if (!require(sf)) install.packages("sf")
if (!require(spdep)) install.packages("spdep")
```


#### Windows

If you use Windows you can install geostan with the following R code:

```{r eval = FALSE}
install.packages("https://connordonegan.github.io/assets/geostan_0.0.1.zip", repos = NULL)
## though some may need:
install.packages("https://connordonegan.github.io/assets/geostan_0.0.1.zip", repos = NULL, type = "binary")
```


#### Linux

If you use Linux you can install the package using:

```{r eval = FALSE}
install.packages("https://connordonegan.github.io/assets/geostan_0.0.1.tar.gz", repos = NULL)
```

or, also for Linux users only, you may install the latest version from source:

```{r eval = FALSE}
remotes::install_github("ConnorDonegan/geostan")
```


#### Mac OS

The package is not currently available for Mac users, sorry.


### Demonstration

As a quick demonstration of the package we will use a simple features (`sf`) dataset with county-level 2016 presidential election results for the state of Ohio. Load the dataset with `geostan` and other helpful packages:

```{r}
library(geostan)
library(sf)
library(ggplot2)
library(rstan)
options(mc.cores = parallel::detectCores())
data(ohio)
## to learn about the data use:
##?ohio
```

We are going to model change in the Republican Presidential vote share from the historic average (2000-2012) to 2016 (`gop_growth`).

```{r}
ggplot(ohio) +
  geom_sf(aes(fill = gop_growth)) +
  scale_fill_gradient2(low = "darkblue",
                       high = "darkred",
                       name = "GOP Growth") +
  theme_void()
```

We can use a Moran plot to visualize the degree of spatial autocorrelation in GOP growth. This plots each observation (on the x-axis) against the mean of neighboring observations (y-axis):

```{r}
W <- shape2mat(ohio, "W")
moran_plot(ohio$gop_growth, W, xlab = "GOP Growth")
```

And we can estimate an ESF model by passing the model formula, data, and spatial connectivity matrix to `stan_esf`:

```{r}
## create binary connectivity matrix
C <- shape2mat(ohio, "B")
fit <- stan_esf(gop_growth ~ 1, data = ohio, C = C)
```

Now `fit` is a list that contains summaries of estimated parameters, the `stanfit` model returned by Stan, diagnostics, and other useful information. We can extract the mean estimate of the spatial filter, which is essentially a spatially varying intercept, with the `spatial` function:

```{r}
sf <- spatial(fit)$mean
```

and visualize it with `ggplot2`:

```{r}
ggplot(ohio) +
  geom_sf(aes(fill = sf)) +
  scale_fill_gradient2(low = "darkblue",
                       high = "darkred",
                       name = "ESF") +
  theme_void()
```

We can also obtain summaries of the residuals and fitted values, and confirm that there is no residual spatial autocorrelation:

```{r}
res <- resid(fit)$mean
f <- fitted(fit)$mean
 ## visualize residual SA
moran_plot(res, W, xlab = "Residuals")
 ## fitted values vs. residuals
ggplot() +
  geom_point(aes(f, res)) +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals") +
  theme_classic()
```

Printing the model returns information about the model call (formula and likelihood function), diagnostics including residual spatial autocorrelation (Moran coefficient) and Widely Applicable Information Criteria (WAIC) for model comparison, plus a summary of the posterior distributions of top-level model parameters. 

```{r}
fit
```

The `Rhat` statistic should always be very near 1 with any divergence from 1 indicating that the chains have not converged to a single distribution. Values further than, say, 1.01 are troubling. The model includes numerous parameters, so the most efficient way to evaluate the diagnostics is to plot them together. To examine autocorrelation in the chains, we can use the estimate of the effective number of posterior samples (`n_eff`):

```{r}
## check effective sample sizes
stan_ess(fit$stanfit) +
  theme_classic()
## check Rhat statistics
stan_rhat(fit$stanfit) +
  theme_classic()
```

We can fit the model the found in Donegan et al. (2020) with a Studen't t likelihood function and spatially lagged covariates:

```{r }
fit3 <- stan_esf(gop_growth ~ historic_gop + log(pop_density) + white_nonhispanic + college_educated + unemployment, 
                 slx = ~ historic_gop + log(pop_density) + college_educated + unemployment,
                 scalex = TRUE, # all X variables scaled to mean = 0, std. dev = 1
                 family = student_t(),
                 data = ohio, 
                 C = C)
```

The spatial connectivity matrix `C` is row-standardized internally before calculating the spatially lagged covariates, so the constructed spatially lagged variables are the mean spatially lagged values.

The `plot` and `print` methods call the bayesplot package, and they accept a character string of parameter names such as `'beta'` for coefficients, `'beta_ev'` for eigenvector coefficients, `'esf'` for the spatial filter. You can also provide more specific names such as `'b_unemployment'`.


```{r fig.width = 6}

plot(fit3, "beta", plotfun = "dens")

```

The `posterior_predict` function can be used for posterior predictive checks together with the bayesplot package (see `?posterior_predict`) and for hypothesis testing. The following example shows how to use it to visualize marginal effects:

```{r }
 # create data.frame all but one of the variables at their mean value (all have mean = 0, std. dev = 1)
newdata <- data.frame(historic_gop = 0, 
                 pop_density = 1, # log(1) = 0
                 college_educated = seq(-2, 2, by = 0.5), 
                 unemployment = 0, 
                 white_nonhispanic = 0
                 )

pred.df = posterior_predict(fit3, # fitted model
                            newdata,  # data for predictions
                            spatial = FALSE, # set the spatial component to zero
                            summary = TRUE, # return a data.frame with 95% credible intervals
)

 # plot the mean of the predictive distribution
fig <- ggplot(pred.df, aes(x=college_educated)) +
    geom_line(aes(y = mu))
  
 # add credible interval around the mean
fig <- fig +
  geom_ribbon(aes(ymin = mu.lwr, 
                  ymax = mu.upr),
              alpha = .5,
              fill = 'darkblue') 
      
 # add cred. interval for the predictive distribution
fig <- fig +
  geom_ribbon(aes(ymin = pred.lwr,
                  ymax = pred.upr),
              alpha = 0.5,
              fill = 'darkblue')
 
fig +
  labs(y = "GOP growth") +
  theme_classic()

```


### Credit

This package incorporates the work of numerous scholars, especially the Stan Development team, Gabry et al.'s **rstantools** package, Morris et al.'s implementation of intrinsic autoregressive models in Stan, and Piironen and Vehtari's regularized horseshoe prior. The Bayesian spatial filtering models implemented here and **geostan** were first introduced in Donegan, Chun, and Hughes (2020) and build on the work of Griffith, Chun, and Li (2019).



Besag, J., York, J., & Mollié, A. (1991). Bayesian image restoration, with two applications in spatial statistics. *Annals of the institute of statistical mathematics*, 43(1), 1-20.

Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M., Guo, J., Li, P., & Riddell, A. (2017). Stan: A probabilistic programming language. *Journal of Statistical Software*, 76(1).

Donegan, C., Chun, Y., & Hughes, A. E. (2020). Bayesian estimation of spatial filters with Moran’s eigenvectors and hierarchical shrinkage priors. *Spatial Statistics* 38: 100450.

Jonah Gabry, Ben Goodrich and Martin Lysy (2020). rstantools: Tools for Developing R Packages Interfacing with 'Stan'. R package version 2.1.1. https://CRAN.R-project.org/package=rstantools
  
Griffith, D., Chun, Y., & Li, B. (2019). *Spatial Regression Analysis Using Eigenvector Spatial Filtering*. Academic Press.

Morris, M., Wheeler-Martin, K., Simpson, D., Mooney, S. J., Gelman, A., & DiMaggio, C. (2019). Bayesian hierarchical spatial models: Implementing the Besag York Mollié model in stan. *Spatial and Spatio-Temporal Epidemiology*, 31, 100301.

Piironen, J., & Vehtari, A. (2017). Sparsity information and regularization in the horseshoe and other shrinkage priors. *Electronic Journal of Statistics*, 11(2), 5018-5051.

Riebler, A., Sørbye, S. H., Simpson, D., & Rue, H. (2016). An intuitive Bayesian spatial model for disease mapping that accounts for scaling. *Statistical Methods in Medical Research*, 25(4), 1145-1165.

Watanabe, S. (2013). A widely applicable Bayesian information criterion. *Journal of Machine Learning Research*, 14(Mar), 867-897.

Wikle, C. K., Berliner, L. M., & Cressie, N. (1998). Hierarchical Bayesian space-time models. *Environmental and Ecological Statistics*, 5(2), 117-154.
